---
title: "kia theft assignment"
author: "Bobga-Herman Gwanvoma"
date: "2025-01-20"
output: pdf_document
---
Load the Data
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)
library(treemap)
library(scales)
library(ggforce)


# Load datasets
vice_news_data <- read_excel("C:/Users/bobi/Documents/DSC 640/w5&6/Motherboard VICE News Kia Hyundai Theft Data.xlsx")
kia_hyundai_thefts <- read.csv("C:/Users/bobi/Documents/DSC 640/w5&6/kiaHyundaiThefts.csv")
car_thefts_map <- read.csv("C:/Users/bobi/Documents/DSC 640/w5&6/carTheftsMap.csv")
milwaukee_data <- read.csv("C:/Users/bobi/Documents/DSC 640/w5&6/KiaHyundaiMilwaukeeData.csv")

# View structure of datasets
str(vice_news_data)
str(kia_hyundai_thefts)
str(car_thefts_map)
str(milwaukee_data)

```
Clean and Preprocess the Data
```{r}
# Clean VICE News Data
vice_clean <- vice_news_data %>%
  select(-starts_with("Unnamed")) %>%  # Remove irrelevant columns
  drop_na()  # Remove rows with missing values

# Clean Kia Hyundai Thefts
kia_hyundai_clean <- kia_hyundai_thefts %>%
  mutate(month_year = paste(month, year, sep = "-")) %>%
  select(city, state, countKiaHyundaiThefts, countOtherThefts, percentKiaHyundai, month_year)

# Clean Car Thefts Map
car_thefts_clean <- car_thefts_map %>%
  rename(geo_location = geo_name) %>%
  mutate(percent_change = as.numeric(percentChange2019to2022)) %>%
  drop_na()

# Clean Milwaukee Data
milwaukee_clean <- milwaukee_data %>%
  mutate(month_year = paste(month, year, sep = "-")) %>%
  select(city, state, countKiaHyundaiThefts, countOtherThefts, percentKiaHyundai, month_year)

```
Visualizations
Visualization 1: Pie Chart (Proportion of Kia vs. Hyundai Thefts)

```{r}
# Aggregate Kia/Hyundai thefts
kia_vs_hyundai <- kia_hyundai_clean %>%
  summarise(total_kia_hyundai = sum(countKiaHyundaiThefts),
            total_other = sum(countOtherThefts))

# Create pie chart
pie_data <- c(kia_vs_hyundai$total_kia_hyundai, kia_vs_hyundai$total_other)
labels <- c("Kia/Hyundai", "Other Vehicles")

pie(pie_data, labels = labels, col = c("skyblue", "orange"), main = "Proportion of Kia/Hyundai vs. Other Vehicle Thefts")

```
Visualization 2: Donut Chart (Thefts by Time Period)
```{r}
library(ggplot2)

# Summarize data for donut chart
time_summary <- milwaukee_clean %>%
  group_by(month_year) %>%
  summarise(total_thefts = sum(countKiaHyundaiThefts))

# Donut chart
ggplot(time_summary, aes(x = 2, y = total_thefts, fill = month_year)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  xlim(1, 2.5) +
  theme_void() +
  ggtitle("Thefts Over Time (Donut Chart)")

```
Visualization 3: Stacked Bar Chart (Thefts by Neighborhood)
```{r}
# Stacked bar chart
kia_hyundai_clean %>%
  group_by(city, state) %>%
  summarise(Kia_Hyundai_Thefts = sum(countKiaHyundaiThefts),
            Other_Thefts = sum(countOtherThefts)) %>%
  gather(key = "Type", value = "Count", -city, -state) %>%
  ggplot(aes(x = city, y = Count, fill = Type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Thefts by Neighborhood", x = "City", y = "Number of Thefts")

```
Visualization 4:Tree Map: Theft Distribution by Vehicle Type
```{r}
# Load necessary library
library(treemap)

# Data summarization
vehicle_summary <- milwaukee_clean %>%
  group_by(city) %>%
  summarise(Total_Thefts = sum(countKiaHyundaiThefts))

# Create treemap
treemap(vehicle_summary,
        index = "city",
        vSize = "Total_Thefts",
        title = "Theft Distribution by Vehicle Type",
        palette = "Blues")

```
Visualization 5: Area Chart (Theft Trends Over Time)
```{r}
# Area chart
ggplot(milwaukee_clean, aes(x = month_year, y = countKiaHyundaiThefts, group = 1)) +
  geom_area(fill = "skyblue", alpha = 0.5) +
  geom_line(color = "blue") +
  theme_minimal() +
  labs(title = "Theft Trends Over Time", x = "Time", y = "Kia/Hyundai Thefts")

```
Visualization 6: Proportion of Thefts by Type and City
```{r}
# Multi-layer pie chart
multi_layer_data <- milwaukee_clean %>%
  group_by(city) %>%
  summarise(Kia_Hyundai_Thefts = sum(countKiaHyundaiThefts),
            Other_Thefts = sum(countOtherThefts)) %>%
  pivot_longer(cols = c(Kia_Hyundai_Thefts, Other_Thefts), names_to = "Type", values_to = "Count")

ggplot(multi_layer_data, aes(x0 = 0, y0 = 0, r0 = ifelse(Type == "Kia_Hyundai_Thefts", 0.6, 1), r = ifelse(Type == "Kia_Hyundai_Thefts", 1, 1.4), amount = Count, fill = city)) +
  geom_arc_bar(stat = "pie") +
  scale_fill_brewer(palette = "Set3") +
  coord_fixed() +
  labs(title = "Proportion of Thefts by Type and City", fill = "City") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))


```



```{r}
ggsave("C:/Users/bobi/Documents/DSC 640/w5&6/pie_chart.png")
ggsave("C:/Users/bobi/Documents/DSC 640/w5&6/donut_chart.png")
ggsave("C:/Users/bobi/Documents/DSC 640/w5&6/stacked_bar_chart.png")
ggsave("C:/Users/bobi/Documents/DSC 640/w5&6/tree_map.png")
ggsave("C:/Users/bobi/Documents/DSC 640/w5&6/area_chart.png")
ggsave("C:/Users/bobi/Documents/DSC 640/w5&6/stacked_area_chart.png")

```

